Bound within Boundaries: How Well Do Protected Areas Match Movement Corridors of Their Most Mobile Protected Species?

### 2020-09-21 Chapter_0/04_Manuscript/Manuscript.tex
Finalized revisions

\noindent 3. Permeability within the boundaries of the KAZA-TFCA was more than
double compared to areas outside it. Furthermore, we observed a five-fold
permeability difference among the five KAZA-TFCA countries. We also showed
that major movement corridors of wild dogs run within the KAZA-TFCA, although
some minor routes remained outside formally protected areas.

\noindent 4. Differences in permeability were mainly related to different
degrees of human activities across regions, and to the presence or absence of
rivers, swamps or open water. The relation between permeability and other
landscape features was less evident.\todo{Editor indicated that the
relationships are correlative and not necessarily causal.}

\noindent 5. \textit{Synthesis and Applications}: In this study, we showed how
pertinent dispersal data of a highly mobile species can be used to empirically
evaluate the adequacy of already-existing or planned protected areas.
Furthermore, we observed regional differences in landscape permeability that
highlight the need for a coordinated effort towards maintaining or restoring
connectivity, especially where transboundary dispersal occurs.\todo{According to
suggestions of reviewer 2}
\end{abstract}

@@

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. Accordingly, preserving and protecting
movement corridors between wildlife reserves has become a task of utmost
\todo{Reviewer one pointed out an awkward use of the word utmost. I have changed
it according to his suggestion.} importance \citep{Doerr.2011, Rudnick.2012},
resulting in an ever-growing number of large and often transboundary protected
areas. While boundaries of such areas are often drawn according to expert
opinion and socio-political needs, subjective assessments have revealed
deficiencies in the past \citep{Clevenger.2002, Pullinger.2010}. Thus, an
empirical evaluation of the adequacy of already-existing or planned protected
areas using pertinent animal movement data is paramount for targeted use of
valuable and scarce conservation funds \citep{Pullinger.2010}.

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales (e.g.
\citealp{Chetkiewicz.2006, Squires.2013, Elliot.2014}). The use of relocation
data has shifted the focus from a structural to a functional view on
connectivity, taking into account how species respond to environmental features
\citep{Taylor.2006, Fattebert.2015}.\todo{As per reviewer 2} Identification of
potential movement corridors typically relies on the estimation of permeability
surfaces, which return the ease or willingness at which the focal species
traverses a specific landscape \citep{Sawyer.2011}. Such surfaces are created
based on species' relative selection strengths \citep{Avgar.2017}\todo{reviewer
1}, which can be quantified using a suite of selection functions
\citep{Zeller.2012}. Specifically, selection strengths \todo{reviewer 1} are
estimated by comparing spatial covariates (e.g. environmental and anthropogenic)
at locations visited by the animal to the same spatial covariates at locations
avilable to the animal \todo{Reviewer 1 pointed out correctly that we should
talk about available locations, rather than randomly selected locations}
\citep{Zeller.2012}. Importantly, selection functions rely on adequate landscape
and relocation data that are representative of the process being studied
\citep{Diniz.2020}. Altough selection during residence and dispersal may
coincide \citep{Fattebert.2015},\todo{second reviewer pointed out that space use
can be alike} it appears that relocation data collected on dispersing
individuals outperforms data collected on resident individuals in the detection
of large-scale movement corridors \citep{Elliot.2014, Diniz.2020}. Nevertheless,
dispersal data is inherently difficult to collect and remains scarce in the
connectivity literature \citep{Vasudev.2015}. As such, most permeability
surfaces upon which movement corridors are identified are created using
relocation data collected on resident individuals. This has likely limited our
ability to meaningfully assess the effectiveness of protected areas in securing
functional connectivity for their protected species.\todo{Reviewer 2 desired to
change from: this introduces - to: this can introduce}

One initiative that aims at restoring and enhancing connectivity across large
scales is the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), which
constitutes the world's largest transfrontier conservation area, spanning over
520'000 km\textsuperscript{2} and five countries (\url{www.kavangozambezi.org}).
While the KAZA-TFCA was originally set to facilitate movements of African
elephants (\textit{Loxodonta africana}; \citealp{Tshipa.2017}), it is also key
to the conservation of other wide-ranging species such as African wild dogs
(\textit{Lycaon pictus}; \citealp{Woodroffe.2012, Cozzi.2020}), lions
(\textit{Panthera leo}; \citealp{Elliot.2014, Cushman.2018}), and cheetahs
(\textit{Acinonyx jubatus}; \citealp{Weise.2017}). To date, however, few studies
have attempted to assess the adequacy of the KAZA-TFCA using relevant global
positioning system (GPS) relocation data of its protected species at large
spatial scales \citep{Elliot.2014, Tshipa.2017}.\todo{as per editor} Thus, how
well the boundaries of the KAZA-TFCA reflect natural movement patterns and
dispersal corridors of its most mobile protected species is virtually unknown.

@@

Here, we collected and analyzed GPS relocation data on 16 dispersing wild dogs
in as many dispersing coalitions from a free-ranging population in northern
Botswana to assess the adequacy of the KAZA-TFCA in securing connectivity. We
estimated relative selection strengths\todo{reviewer 1} towards environmental
and anthropogenic landscape features, and used the obtained coefficients to
predict a permeability surface spanning the entire KAZA-TFCA. We then
investigated how landscape permeability varied regionally and internationally,
and we compared permeability within and outside the KAZA-TFCA boundaries.
Finally, we calculated least-cost paths and corridors to identify major movement
routes and to verify that these were successfully covered by the KAZA-TFCA.

@@

\subsection{GPS Relocation Data}
We used a population of free-ranging African wild dogs inhabiting the Okavango
Delta in northern Botswana as a source population for dispersing individuals.
This population has been extensively studied since 1989 \citep{McNutt.1996,
Cozzi.2013, Cozzi.2020, Behr.2020}. Between 2011 and 2019, we systematically
collected GPS relocation data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions). Candidate dispersing individuals were
identified based on age, number of same‐sex siblings, pack size, and presence of
unrelated individuals of the opposite sex in their pack \citep{McNutt.1996,
Behr.2020}\todo{The editor desired to include criteria}. Selected individuals
were immobilized according to protocols described in \cite{Osofsky.1996}, and
fitted with GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic
Aerospace GmbH, Berlin, Germany}) while still with their natal pack. All
required procedures were undertaken and supervised by a Botswana-registered
wildlife veterinarian. Collars produced a tag weight of 300 gramms, accounting
for 1.2\% of a wild dog's average body weight. They were attached using leather
belts that included a drop off mechanism, triggered by a decomposing piece of
cloth.\todo{Editor desired more information on GPS collar} During dispersal, GPS
collars were programmed to record a GPS relocation every 4 hours and to
regularly transmit data via Iridium satellite system to a base station.

Because we were interested in dispersal behavior only, we discarded any GPS data
collected while individuals were still with their natal packs and after
settlement in a new territory \citep{Cozzi.2020}. We identified the exact time
of emigration and settlement based on direct field observations and through
visual inspection of the net squared displacement (NSD) metric. NSD quantifies
the squared \todo{Reviewer 1 pointed out correctly that we need to talk about
the \textit{squared} euclidean distance} Euclidean distance of a relocation to a
reference point \citep{Borger.2012}, which in our case was the center of the
dispersing coalition's natal home range. Thus, dispersal was deemed to have
started when a coalition had left its natal home range and continued until the
NSD metric remained stationary, indicating that the coalition had successfully
settled (Figure S1). In our analysis, we did not differentiate between male and
female dispersing coalitions, for previous research found little differences
between sexes during dispersal \citep{Woodroffe.2019, Cozzi.2020}.

\subsection{Spatial Covariates}
To investigate relative selection strengths \todo{As per reviewer 1} of
dispersing wild dogs, we used a set of geo-referenced covariates
(\Cref{Covariates}) that we aggregated in the categories \textit{land cover},
\textit{protection status}, and \textit{human influence}. Land cover comprised
the covariates water cover, distance to water, percentage cover by
shrubs/grassland, and percentage cover by trees. To capture the pulsing behavior
of the Okavango Delta, we frequently updated layers for water cover and
corresponding layers depicting distance to water. Protection status contained a
binary indicator of whether an area was protected or not. Human influence
included covariates rendering human density, the presence of roads, and the
distance to roads.\todo{Reviewer 2 found the description of the covariates
unclear} We prepared all covariates as spatial raster layers from freely
available online services and from remotely sensed satellite imagery. To ensure
a consistent resolution (i.e. cell-size or grain) across covariates, we
coarsened or interpolated all layers to a resolution of 250m x 250m. For further
details on the preparation of each covariate and their sources, see Appendix
A.3. We performed processing and manipulation of data as well as all spatial and
statistical analyses using R, version 3.6.1 \citep{R.2019}.

@@

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' relative selection strentghs towards \todo{Reviewer 1
suggested to use relative selection strength, rather than preference} the
above-mentioned spatial covariates. That is, we paired each realized step (i.e.
the connecting line between two consecutive GPS relocations;
\citealp{Turchin.1998}) with 24 random steps that were generated by sampling
turning angles from a uniform distribution U(\(-\pi, +\pi\)) and step lengths
from a gamma distribution fitted to realized steps \citep{Avgar.2016}. A
realized step and its 24 associated random steps formed a stratum and received a
unique identifier. Along each step, we extracted the above-mentioned covariates
(Table S3), standardized extracted values using a z-score transformation, and
checked for correlation using Pearson's Correlation Coefficient \(r\). None of
the covariates were overly correlated (\(|r| > 0.6\); \citealp{Latham.2011}) and
we retained all of them for modeling. Our habitat selection model then assumed
that dispersing wild dogs assigned a selection score \(w(x)\) of the following
exponential form to each realized and random step \citep{Fortin.2005}:

@@

\noindent The selection score \(w(x)\) of a step depended on its associated
covariates (\(x_1, x_2, ..., x_n\)), as well as on the animal's relative
selection strengths \todo{As per reviewer 1} towards these covariates
(\(\beta_1, \beta_2, ..., \beta_n\)). To estimate relative selection strengths
\todo{As per reviewer 1} (i.e. the \(\beta\)'s) for each covariate, we used
mixed effects conditional logistic regression analysis as suggested by
\cite{Muff.2020}. We implemented their method using the R-package
\textit{glmmTMB} \citep{Mollie.2017} and used dispersing coalition ID to model
random slopes. We also modelled random intercepts with an arbitrary high
variance of 10\textsuperscript{6} to make use of the poisson trick (see
\citealp{Muff.2020}). We defined three movement metrics, namely the cosine of
the turning angle (\(cos(ta)\)), the step length (\(sl\)) and the natural
logarithm of the step length (\(log(sl)\)), as core covariates and ran stepwise
forward model selection based on Akaike's Information Criterion (AIC;
\citealp{Burnham.2002}) for all other covariates.\todo{Reviewer 1 asked for more
details in the random effects. I therefore decided to update this section too}
The inclusion of movement metrics served to reduce biases in estimated habitat
selection coefficients \todo{as per revierwer 1} that may have arisen due to
movement behavior \citep{Avgar.2016}. To validate the predictive power of the
most parsimonious habitat selection model, we ran k-fold cross-validation for
case-control studies as described in \cite{Fortin.2009} (details in Appendix
A.5).

\subsection{Permeability Surface}
Using the most parsimonious habitat selection model, we predicted a permeability
surface spanning the entire extent of the KAZA-TFCA. That is, we applied
\ref{EQ1} to our spatial covariates and calculated the selection score \(w(x)\)
for each raster cell. Because our representation of water was dynamic (to
properly render the pulsing behavior of the Okavango Delta) we collapsed all
dynamic water maps into a single static map using areas that were covered by
water in at least 10\% of the cases. Using the resulting static map we also
calculated a layer returning the distance to water. Because the extent of the
delta covers merely 6\% of the entire KAZA-TFCA, the use of static maps should
introduce little bias in predicted permeability. To reduce the influence of
outliers in predicted permeability scores, we followed \cite{Squires.2013} and
curtailed predicted scores between the 1\textsuperscript{st} and
99\textsuperscript{th} percentile of their original values. To compare
permeability across different regions, we normalized the permeability surface to
a range between 0 (most impermeable) and 1 (most permeable), and we determined
median permeability within and outside the KAZA-TFCA, within and outside
formally protected areas, and within each of the five KAZA-TFCA countries.

\subsection{Least-Cost Paths and Corridors}
To identify movement corridors of dispersing wild dogs, we specified source
points and calculated factorial least-cost paths (LCPs) as well as factorial
least-cost corridors (LCCs) among them \citep{Elliot.2014}. We generated source
points by overlaying the study area with a regular grid of points spaced at 100
km. We only considered those points that fell within protected areas \(>\) 700
km\textsuperscript{2}, which conforms with home-range requirements of African
wild dogs \citep{Pomilia.2015}. Finally, we defined centroids as source points
for those protected areas \(>\) 700 km\textsuperscript{2} that were not assigned
any source points from the regular grid. Because wild dogs residing outside of
protected areas are rare and likely unviable \citep{VanDerMeer.2014}, we
consider our selection of source points within protected areas to be
appropriate. In total, we generated 68 source points, which resulted in 2'278
unique pairwise combinations and therefore 2'278 unique LCPs and LCCs. We
computed factorial LCPs and LCCs between source points using the R-package
\textit{gdistance} (further details in Appendix A.7). After computation, we
tallied overlapping LCPs and LCCs, respectively, into single connectivity maps.

\section{Results}
\subsection{Dispersal Events}
In total, we collected 4'169 GPS relocations during dispersal (Figure S2 \&
Table S1), resulting in an average of 261 (\(SD\) = 207) locations per
dispersing coalition. Coalitions on average dispersed for 48 days (\(SD\) = 44),
covered a mean Euclidean distance of 54 km (\(SD\) = 71) and a cumulative
distance of 597 km (\(SD\) = 508). One female coalition dispersed far east into
the Hwange Hwange National Park and covered a cumulative distance of over 360 km
in under 9 days.\todo{Editor desired a quick summary of the dispersal data.}

\subsection{Habitat Selection Model}
Our most parsimonious habitat selection model (\(\Delta AIC > 2\) than any
alternative model; Table S4) retained the covariates \textit{water},
\textit{distance to water}, \textit{trees}, \textit{shrubs/grassland}, and
\textit{human influence}, beside the fixed covariates \textit{cos(ta)},
\textit{sl}, and \textit{log(sl)} (\Cref{PermeabilityResults}a). The positive
coefficients for \(sl\) (\(\beta\) = 0.03, 95\% CI = -0.09 to 0.15) and
\(log(sl)\) (\(\beta\) = 0.12, 95\% CI = 0.02 to 0.22) indicated that our
initial (selection-biased) gamma distribution produced step-lengths slightly
shorter than a selection-free gamma distribution would. Similarly, the positive
coefficient for \(cos(ta)\) (\(\beta\) = 0.10; 95\% CI = 0.05 to 0.15) suggested
that our uniform distribution for turning angles produced turning angles with
too little directionality. \todo{Reviewer 1 nicely explained that the
coefficients for log(sl) and cos(ta) imply that the initial distributions were
selection-biased} With respect to environmental covariates, dispersers avoided
moving through water (\(\beta\) = -0.53, 95\% CI -0.79 to -0.27) but selected
for locations in its vicinity, although the latter effect was not significant
(\(\beta\) = -0.33, 95\% CI = -0.73 to 0.08). Dispersers avoided areas that were
densely covered by trees (\(\beta\) = -0.31, CI = -0.47 to -0.15) and preferred
areas covered by shrubs/grassland (\(\beta\) = 0.25, 95\% CI = 0.07 to 0.42).
Finally, dispersers avoided areas that were influenced by humans (\(\beta\) =
-0.45, 95\% CI = -0.82 to -0.08). With the exepction of \textit{distance to
water} (\(SD_{RE}\) = 0.57), random effects revealed little variability between
different dispersal coalitions' selection strengths (i.e. all with \(SD_{RE}
\leq 0.22\), Figure S7).\todo{Reviewer 1 desired more information on the random
effects}

@@

Our results emphasize that human influences constitute some of the main barriers
to connectivity among wild dog populations. According to our model, dispersers
avoided corssing human dominated landscapes whenever given the
choice.\todo{According to reviewer 1 we should point out that our model compares
available vs. unavailable} This conforms to findings on dispersing wild dogs
from eastern Africa \citep{Masenga.2016, Oneill.2020} but conflicts with
findings from South Africa by \cite{DaviesMostert.2012}, who reported a high
willingness of dispersers to cross human-dominated landscapes. We believe that
such differences are due to the unavailability of alternative routes through
natural landscapes, which may have forced dispersers in South Africa to cross
human dominated landscapes despite a strong aversion to do so. In this regard,
our representation of dispersal corridors and the resulting connectivity appear
conservative, as dispersers may be able to make the best out of a bad situation
and cross landscapes characterized by considerably unfavorable conditions
\citep{Palomares.2000, Elliot.2014}. Unfortunately, \cite{DaviesMostert.2012}
only recorded net dispersal distances, thereby precluding a more indepth
understanding of the impact of humans on dispersal behavior.\todo{As per
reviewer 2} Nevertheless, successful conservation of this species relies on
policymakers' and local authorities' willingness and ability to provide and
conserve natural areas that remain free from anthropogenic pressures. This is
not only paramount in light of increasing connectivity and facilitating
dispersal, but also in terms of reducing human-caused mortality during
dispersal. In fact, previous studies have shown that human-caused mortality
represents a major threat to wild dogs' ability to disperse
\citep{Woodroffe.2019, Cozzi.2020}.

@@

Although dispersers avoided moving through water, they selected locations in its
vicinity. This behavior may be caused by the occurence of prey close to water
\citep{Bonyongo.2005}. For the same reason, however, competitors such as lions,
spotted hyenas, and resident wild dogs may also use areas close to water
\citep{Valeix.2010}, thereby occasionally forcing dispersing wild dogs to move
into prey-poorer areas away from water \citep{Creel.2002, Mills.1997}. This
reasoning may also explain why the effect of distance to water was
insignificant.\todo{Reviewer 1 suggested to add this} Given the influence that
resident conspecifics, competitors, and prey can have on dispersers
\citep{Cozzi.2018, Armansin.2019} future studies should strive to collect and
incorporate intra- and interspecific relationships into analyses of landscape
connectivity.

Overall, our findings on habitat selection during dispersal coincide with
findings from dipsersing wild dogs in Kenya \citep{Oneill.2020}, suggesting that
there are strong commonalities dispersers from these very different ecosystems.
Thus, despite wild dogs' ability to cope with diverse habitats and adapt to
changing conditions \citep{Woodroffe.2011}, the fundamental factors included in
our study appear to influence wild dogs from other ecosystems alike.
Nevertheless, expanding our analysis to dispersers emigrating from other source
populations would invaluably contribute to our understanding of
dispersal.\todo{Editor desired info on applicability}

>>>>

\noindent 3. Permeability within the boundaries of the KAZA-TFCA was more than
double compared to areas outside it. Furthermore, we observed a five-fold
permeability difference among the five KAZA-TFCA countries. We also showed
that major movement corridors of wild dogs ran within the KAZA-TFCA, although
some minor routes remained outside formally protected areas.

\noindent 4. Differences in permeability were mainly related to different
degrees of human activities across regions, and to the presence or absence of
rivers, swamps or open water. The relation between permeability and other
landscape features was less evident.

\noindent 5. \textit{Synthesis and Applications}: In this study, we showed how
pertinent dispersal data of a highly mobile species can be used to empirically
evaluate the adequacy of already-existing or planned protected areas.
Furthermore, we observed regional differences in landscape permeability that
highlight the need for a coordinated effort towards maintaining or restoring
connectivity, especially where transboundary dispersal occurs.
\end{abstract}

@@

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. Accordingly, preserving and protecting
movement corridors between wildlife reserves has become a task of utmost
importance \citep{Doerr.2011, Rudnick.2012}, resulting in an ever-growing number
of large and often transboundary protected areas. While boundaries of such areas
are often drawn according to expert opinion and socio-political needs,
subjective assessments have revealed deficiencies in the past
\citep{Clevenger.2002, Pullinger.2010}. Thus, an empirical evaluation of the
adequacy of already-existing or planned protected areas using pertinent animal
movement data is paramount for targeted use of valuable and scarce conservation
funds \citep{Pullinger.2010}.

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales (e.g.
\citealp{Chetkiewicz.2006, Squires.2013, Elliot.2014}). Identification of
potential movement corridors typically relies on the estimation of permeability
surfaces, which return the ease or willingness at which the focal species
traverses a specific landscape \citep{Sawyer.2011}. Such surfaces are created
based on species' relative selection strengths \citep{Avgar.2017}, which can be
quantified using a suite of selection functions \citep{Zeller.2012}.
Specifically, selection strengths are estimated by comparing spatial covariates
(e.g. environmental and anthropogenic) at locations visited by the animal to the
same spatial covariates at locations avilable to the animal \citep{Zeller.2012}.
Importantly, selection functions require adequate landscape and relocation data
that are representative of the process being studied \citep{Diniz.2020}. Altough
selection during residence and dispersal may coincide \citep{Fattebert.2015}, it
appears that relocation data collected on dispersing individuals outperforms
data collected on resident individuals in the detection of large-scale movement
corridors \citep{Elliot.2014, Abrahms.2017, Diniz.2020}. Nevertheless, dispersal
data is inherently difficult to collect and remains scarce in the connectivity
literature \citep{Vasudev.2015}. As such, most permeability surfaces upon which
movement corridors are identified are created using relocation data collected on
resident individuals. This has likely limited our ability to meaningfully assess
the effectiveness of protected areas in securing functional connectivity for
their protected species.

One initiative that aims at restoring and enhancing connectivity across large
scales is the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), which
constitutes the world's largest transfrontier conservation area, spanning over
520'000 km\textsuperscript{2} and five countries (\url{www.kavangozambezi.org}).
While the KAZA-TFCA was originally set to facilitate movements of African
elephants (\textit{Loxodonta africana}; \citealp{Tshipa.2017}), it is also key
to the conservation of other wide-ranging species such as African wild dogs
(\textit{Lycaon pictus}; \citealp{Woodroffe.2012, Cozzi.2020}), lions
(\textit{Panthera leo}; \citealp{Elliot.2014, Cushman.2018}), and cheetahs
(\textit{Acinonyx jubatus}; \citealp{Weise.2017}). To date, however, few studies
have attempted to assess the adequacy of the KAZA-TFCA using global positioning
system (GPS) relocation data of its protected species at large spatial scales
\citep{Elliot.2014, Tshipa.2017, Brennan.2020}. Thus, how well the boundaries of
the KAZA-TFCA reflect natural movement patterns and dispersal corridors of its
most mobile protected species is virtually unknown.

@@

Here, we collected and analyzed GPS relocation data on 16 dispersing wild dogs
in as many dispersing coalitions from a free-ranging population in northern
Botswana to assess the adequacy of the KAZA-TFCA in securing connectivity. We
estimated relative selection strengths towards environmental and anthropogenic
landscape features, and used the obtained coefficients to predict a permeability
surface spanning the entire KAZA-TFCA. We then investigated how landscape
permeability varied regionally and internationally, and we compared permeability
within and outside the KAZA-TFCA boundaries. Finally, we calculated least-cost
paths and corridors to identify major movement routes and to verify that these
were successfully covered by the KAZA-TFCA.

@@

\subsection{GPS Relocation Data}
We used a population of free-ranging African wild dogs inhabiting the Okavango
Delta in northern Botswana as a source population for dispersing individuals.
This population has been extensively studied since 1989 \citep{McNutt.1996,
Cozzi.2013, Cozzi.2020, Behr.2020}. Between 2011 and 2019, we systematically
collected GPS relocation data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions). Candidate dispersing individuals were
identified based on age, number of same‐sex siblings, pack size, and presence of
unrelated individuals of the opposite sex in their pack \citep{McNutt.1996,
Behr.2020}. Selected individuals were immobilized according to protocols
described in \cite{Osofsky.1996}, and fitted with GPS/Satellite radio collars
(\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin, Germany}) while still
with their natal pack. Collars weighed 330g, accounting for about 1.5\% of a
wild dog's body weight. A 5cm long decomposable cotton piece was added to the
collar belt to guarantee collar drop-off after about 12-18 months. All required
procedures were undertaken and supervised by a Botswana-registered wildlife
veterinarian. During dispersal, GPS collars were programmed to record a GPS
relocation every 4 hours and to regularly transmit data via iridium satellite
system to a base station.

Because we were interested in dispersal behavior only, we discarded any GPS data
collected while individuals were still with their natal packs and after
settlement in a new territory \citep{Cozzi.2020}. We identified the exact time
of emigration and settlement based on direct field observations and through
visual inspection of the net squared displacement (NSD) metric. NSD quantifies
the squared Euclidean distance of a relocation to a reference point
\citep{Borger.2012}, which in our case was the center of the dispersing
coalition's natal home range. Thus, dispersal was deemed to have started when a
coalition had left its natal home range and continued until the NSD metric
remained stationary, indicating that the coalition had successfully settled
(Figure S1). In our analysis, we did not differentiate between male and female
dispersing coalitions, for previous research found little differences between
sexes during dispersal \citep{Woodroffe.2019, Cozzi.2020}.

\subsection{Spatial Covariates}
To investigate relative selection strengths of dispersing wild dogs, we used a
set of geo-referenced covariates (\Cref{Covariates}) that we aggregated in the
categories \textit{land cover}, \textit{protection status}, and
\textit{anthropogenic}. \textit{Land cover} comprised the covariates water cover
(binary), distance to water (continuous), percentage cover by shrubs/grassland
(continuous), and percentage cover by trees (continuous). To capture the pulsing
behavior of the Okavango Delta, we classified satellite imagery and frequently
updated layers for water cover and corresponding layers depicting distance to
water. \textit{Protection status} contained a binary covariate, indicating
whether an area was protected or not. \textit{Anthropogenic} included covariates
rendering the presence of roads (binary), the distance to roads (continuous),
and a proxy for human influence (continuous) that took into account human
density, farming, and roads. We prepared all covariates as spatial raster layers
from freely available online services and from remotely sensed satellite
imagery. To ensure a consistent resolution (i.e. cell-size or grain) across
covariates, we coarsened or interpolated all layers to a resolution of 250m x
250m. For further details on the preparation and source of each covariate, see
Appendix A.3. We performed processing and manipulation of data as well as all
spatial and statistical analyses using R, version 3.6.1 \citep{R.2019}.

@@

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' relative selection strentghs towards the above-mentioned
spatial covariates. That is, we paired each realized step (i.e. the connecting
line between two consecutive GPS relocations; \citealp{Turchin.1998}) with 24
random steps that were generated by sampling turning angles from a uniform
distribution U(\(-\pi, +\pi\)) and step lengths from a gamma distribution fitted
to realized steps \citep{Avgar.2016}. A realized step and its 24 associated
random steps formed a stratum and received a unique identifier. Along each step,
we extracted the above-mentioned covariates (Table S3), standardized extracted
values using a z-score transformation, and checked for correlation using
Pearson's Correlation Coefficient \(r\). None of the covariates were overly
correlated (\(|r| > 0.6\); \citealp{Latham.2011}) and we retained all of them
for modeling. Our habitat selection model then assumed that dispersing wild dogs
assigned a selection score \(w(x)\) of the following exponential form to each
realized and random step \citep{Fortin.2005}:

@@

\noindent The selection score \(w(x)\) of a step depended on its associated
covariates (\(x_1, x_2, ..., x_n\)), as well as on the animal's relative
selection strengths towards these covariates (\(\beta_1, \beta_2, ...,
\beta_n\)). To estimate relative selection strengths (i.e. the \(\beta\)'s) for
each covariate, we used mixed effects conditional logistic regression analysis
as suggested by \cite{Muff.2020}. We implemented their method using the
R-package \textit{glmmTMB} \citep{Mollie.2017} and used dispersing coalition ID
to model random slopes. We also modelled random intercepts with an arbitrary
high variance of 10\textsuperscript{6} to make use of the poisson trick (see
\citealp{Muff.2020}). We defined three movement metrics, namely the cosine of
the turning angle (\(cos(ta)\)), the step length (\(sl\)) and the logarithm of
the step length (\(log(sl)\)), as core covariates and ran stepwise forward model
selection based on Akaike's Information Criterion (AIC; \citealp{Burnham.2002})
for all other covariates. The inclusion of movement metrics served to reduce
biases in estimated habitat selection coefficients that may have arisen due to
movement behavior \citep{Avgar.2016}. To validate the predictive power of the
most parsimonious habitat selection model, we ran k-fold cross-validation for
case-control studies as described in \cite{Fortin.2009} (details in Appendix
A.5).

\subsection{Permeability Surface}
Using the most parsimonious habitat selection model, we predicted a permeability
surface spanning the entire extent of the KAZA-TFCA. That is, we applied
\ref{EQ1} to our spatial covariates and calculated the selection score \(w(x)\)
for each raster cell. Because our representation of water was dynamic we
collapsed all dynamic water maps into a single map using areas that were covered
by water in at least 10\% of the cases. We used the resulting map to also
calculate a layer returning the distance to water. To reduce the influence of
outliers in predicted permeability scores, we followed \cite{Squires.2013} and
curtailed predicted scores between the 1\textsuperscript{st} and
99\textsuperscript{th} percentile of their original values. To compare
permeability across different regions, we normalized the permeability surface to
a range between 0 (most impermeable) and 1 (most permeable), and we determined
median permeability within and outside the KAZA-TFCA, within and outside
formally protected areas, and within each of the five KAZA-TFCA countries.

\subsection{Least-Cost Paths and Corridors}
To identify movement corridors of dispersing wild dogs, we specified source
points and calculated factorial least-cost paths (LCPs) as well as factorial
least-cost corridors (LCCs) among them \citep{Elliot.2014}. We generated source
points by overlaying the study area with a regular grid of points spaced at 100
km. We only considered those points that fell within protected areas \(>\) 700
km\textsuperscript{2}, which conforms with home-range requirements of African
wild dogs \citep{Pomilia.2015}. Finally, we defined centroids as source points
for those protected areas \(>\) 700 km\textsuperscript{2} that were not assigned
any source points from the regular grid. Because wild dogs residing outside of
protected areas are rare and unviable \citep{VanDerMeer.2014}, we consider our
selection of source points within protected areas to be appropriate. In total,
we generated 68 source points, which resulted in 2'278 unique pairwise
combinations and therefore 2'278 unique LCPs and LCCs. We computed factorial
LCPs and LCCs between source points using the R-package \textit{gdistance}
(further details in Appendix A.7). After computation, we tallied overlapping
LCPs and LCCs, respectively, into single connectivity maps. Because LCPs return
discrete paths, whereas LCCs return continuous corridors, we present both
methods.

\section{Results}
\subsection{Dispersal Events}
In total, we collected 4'169 GPS relocations during dispersal (Figure S2 \&
Table S1), resulting in an average of 261 (\(SD\) = 207) locations per
dispersing coalition. Coalitions on average dispersed for 48 days (\(SD\) = 44),
covered a mean Euclidean distance of 54 km (\(SD\) = 71) and a cumulative
distance of 597 km (\(SD\) = 508).

\subsection{Habitat Selection Model}
Our most parsimonious habitat selection model (\(\Delta AIC > 2\) than any
alternative model; Table S4) retained the covariates \textit{water},
\textit{distance to water}, \textit{trees}, \textit{shrubs/grassland}, and
\textit{human influence}, beside the fixed covariates \textit{cos(ta)},
\textit{sl}, and \textit{log(sl)} (\Cref{PermeabilityResults}a). With respect to
environmental covariates, dispersers avoided moving through water (\(\beta\) =
-0.53, 95\% CI -0.79 to -0.27) but selected for locations in its vicinity,
although the latter effect was not significant (\(\beta\) = -0.33, 95\% CI =
-0.73 to 0.08). Dispersers avoided areas that were densely covered by trees
(\(\beta\) = -0.31, CI = -0.47 to -0.15) and preferred areas covered by
shrubs/grassland (\(\beta\) = 0.25, 95\% CI = 0.07 to 0.42). Finally, dispersers
avoided areas that were influenced by humans (\(\beta\) = -0.45, 95\% CI = -0.82
to -0.08). With the exepction of \textit{distance to water}, random effects
revealed little variability between dispersal coalitions (Appendix A6).

% The positive coefficient for \(cos(ta)\) (\(\beta\) = 0.10; 95\% CI = 0.05 to
% 0.15) suggested that the uniform distribution for turning angles produced too
% little directionality. Similarly, the positive coefficient for \(log(sl)\)
% (\(\beta\) = 0.12, 95\% CI = 0.02 to 0.22) indicated that our fitted gamma
% distribution produced step-lengths slightly shorter than the true
% selection-free gamma distribution would have. The estimate for \(sl\), on the
% other hand, was insignificant (\(\beta\) = 0.03, 95\% CI = -0.09 to 0.15).

@@

Our results emphasize that human influences constitute some of the main barriers
to connectivity among wild dog populations. This conforms to findings on
dispersing wild dogs from eastern Africa \citep{Masenga.2016, Oneill.2020} but
conflicts with findings from South Africa by \cite{DaviesMostert.2012}, who
reported a high willingness of dispersers to cross human-dominated landscapes.
Such differences may arise from the fact that our model infers preferences by
comparing \textit{used} and \textit{available} habitats, whereas
\cite{DaviesMostert.2012} only recorded net dispersal distances, thereby
precluding such an analysis. Thus, we believe that differences to
\cite{DaviesMostert.2012} may be explained by the unavailability of alternative
routes through natural landscapes, which may have forced dispersers in South
Africa to cross human dominated landscapes despite a strong aversion to do so.
In this regard, our representation of dispersal corridors and the resulting
connectivity appear conservative, as dispersers may be able to make the best out
of a bad situation and cross landscapes characterized by considerably
unfavorable conditions \citep{Palomares.2000, Elliot.2014}. Nevertheless,
successful conservation of this species relies on policymakers' and local
authorities' willingness and ability to provide and conserve natural areas that
remain free from anthropogenic pressures. This is not only paramount in light of
increasing connectivity and facilitating dispersal, but also in terms of
reducing human-caused mortality during dispersal. In fact, previous studies have
shown that human-caused mortality represents a major threat to wild dogs'
ability to disperse \citep{Woodroffe.2019, Cozzi.2020}.

@@

Although dispersers avoided moving through water, they selected locations in its
vicinity. This behavior may be caused by the occurence of prey close to water
\citep{Bonyongo.2005}. For the same reason, however, competitors such as lions,
spotted hyenas, and resident wild dogs may also use areas close to water
\citep{Valeix.2010}, thereby occasionally forcing dispersing wild dogs to switch
behavior and move into prey-poorer areas away from water \citep{Creel.2002,
Mills.1997}. This reasoning may also explain the large confidence intervals
surrounding the corresponding \(\beta\)-estimate. Given the influence that
resident conspecifics, competitors, and prey can have on dispersers
\citep{Cozzi.2018, Armansin.2019} future studies should strive to collect and
incorporate intra- and interspecific relationships into analyses of landscape
connectivity.

Overall, our findings on habitat selection during dispersal coincide with
findings from dipsersing wild dogs in Kenya \citep{Oneill.2020}, suggesting that
there are strong commonalities between dispersers from these very different
ecosystems. Thus, despite wild dogs' ability to cope with diverse habitats and
adapt to changing conditions \citep{Woodroffe.2011}, the fundamental factors
included in our study appear to influence wild dogs from other ecosystems alike.
Nevertheless, expanding our analysis to dispersers emigrating from other source
populations would invaluably contribute to our understanding of dispersal.
